---
title: "Study of weather harmful events in US (1950-2011)"
author: "José Manuel Peñas"
date: "18 de mayo de 2018"
output: html_document
---
## Synopsis





## Data Processing

From the US **National Weather Service** I have got a file with data recorded about the weather events, storms, since **1950** to November of **2011**.  
The data for this assignment come in the form of a comma-separated-value file compressed via the **bzip2** algorithm to reduce its size. You can download the file from the course web [site](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).  
The information about the variables of the file can be consulted on the following links:

- National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf) 

### KNTIR Setup
First, the **knitr** preferences and required libraries for the next r code chunks are set.
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
```

```{r cars}
summary(cars)
```

## Results

You can also embed plots, for example:

```{r pressure}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
